European or American? A re-examination of the transatlantic influence over the IMF

### 2025-07-28 manuscript/manuscript.qmd
europe lit revie

Qualitative approaches to this issue are often hampered by the Fund’s closed-door, consensus-based decision-making [@stone2011, pp.55-56]—precisely the kind of opaque process exemplified by the “gentlemen’s agreement.” On the one hand, the prevalence of informal conventions limits the utility of formal institutional analyses (e.g., voting power or veto rights). On the other, the secretive nature of negotiations means scholars often observe the outcomes, but not the underlying power struggles. While case studies such as Mexico’s bailout or the European debt crisis [@lipscy2018; @copelovitch2010a; @hodson2014] provide compelling narratives, they lack transatlantic comparability and generalisability.

@@

International Relations theory, particularly realism, has long emphasised that international organisations reflect underlying distributions of power rather than operating as neutral or autonomous actors. As Mearsheimer famously argued, international institutions are “basically a reflection of the distribution of power in the world” [@mearsheimer1994]. The International Monetary Fund (IMF), despite its formal presentation as a rules-based and technocratic institution [@barnett2004; @chwieroth2015; @lutz2019], is no exception [@copelovitch2010a; @stone2011; @vreeland2009]. Empirical studies have consistently shown that geopolitical considerations play a central role in IMF lending decisions.

@@

For the strand of literature that examines American and European influence separately, nearly all studies identify the United States as the dominant actor within the IMF [@stone2011, p.51]. Qualitative accounts emphasise the United States’ foundational role in shaping the IMF’s governance structure [@stone2011], the Fund’s practice of making direct enquiries with Washington for major lending decisions [@southard1979; @woods2003], and its unique status as the only country-level veto player, stemming from its position as the largest financial contributor [@woods2003; @stone2011]. The US also benefits from significant informational advantages over other members [@stone2011, p.57] and proximity-based access, with the IMF headquartered in Washington, D.C. [@stone2011, p.57]. Numerous case studies further demonstrate instances of US political pressure resulting in lenient loan conditions for key geopolitical allies, including Zaire, Egypt, Mexico, Argentina, and South Korea [@vreeland2009; @copelovitch2010a; @lipscy2018]. This pattern of influence is consistent with the United States’ broader role across international financial institutions, particularly the World Bank, where American dominance is widely acknowledged [@wade2002; @woods2003; @clark2020].

In comparison, the literature on European influence is notably smaller. The most visible manifestation of European power within the Fund is the tradition of European leadership: all Managing Directors of the IMF since its inception have been European nationals [@kahler2001]. Scholars have also pointed to Europe-backed lending decisions, such as those during the Eurozone crisis (notably Greece) [@lipscy2018; @hodson2014], and loans to former colonies in Africa [@stone2004]. However, while these studies offer important insights, they tend to remain case-specific or anecdotal. Crucially, they stop short of offering a systematic comparison between American and European influence, leaving open the question of how these two centres of power operate within the IMF.

In principle, quantitative approaches are well suited for directly comparing the influence of different great powers within international organisations. In the IMF context, however, most such studies focus on American influence, with little systematic attention paid to European power. These studies typically proxy the strategic interests of major powers using indicators such as UN voting alignment [@thacker1999; @dreher2009], bilateral trade intensity [@barro2005], and financial exposure [@oatley2004]. Yet, much of this literature has remained US-centric, often using only American alignment as the independent variable of interest [@thacker1999; @oatley2004; @dreher2009]. Other studies pool European indicators with US ones under the broader category of “Western” influence, thus masking potential differences between the two [@lipscy2018; @barro2005]. These ties have been shown to correlate with more favourable IMF lending outcomes—larger loans, higher approval rates, and fewer or less stringent conditions [@lipscy2018; @barro2005; @dreher2009]. Yet, despite the availability of these tools, comparative assessments between the US and Europe remain largely absent from this literature.

This paper addresses this gap by conducting a disaggregated, quantitative comparison of US and European influence on IMF lending outcomes. By distinguishing between the two major transatlantic actors, we provide new insights into the internal dynamics of power within the IMF and contribute to broader debates over institutional legitimacy, hegemony, and the geopolitics of economic governance.

\color{red}
However, empirical studies suggested that countries with low international reserves, high current account deficits, and low GDP growth are more likely to receive IMF funding assistance [@barro2005; @knight1997], despite such indicators often reflecting a limited capacity to repay.
\color{black}

>>>>

Qualitative approaches to this issue are often hampered by the Fund’s closed-door, consensus-based decision-making—precisely the kind of opaque process exemplified by the “gentlemen’s agreement.” On the one hand, the prevalence of informal conventions limits the utility of formal institutional analyses (e.g., voting power or veto rights). On the other, the secretive nature of negotiations means scholars often observe the outcomes, but not the underlying power struggles. While case studies such as Mexico’s bailout or the European debt crisis [@lipscy2018; @copelovitch2010a; @hodson2014] provide compelling narratives, they lack transatlantic comparability and generalisability.

@@

International Relations theory, particularly realism, has long emphasised that international organisations reflect underlying distributions of power rather than operating as neutral or autonomous actors. As Mearsheimer famously argued, international institutions are “basically a reflection of the distribution of power in the world” [@mearsheimer1994]. The International Monetary Fund (IMF), despite its formal presentation as a rules-based and technocratic institution [@barnett2004; @chwieroth2015; @lutz2019], is no exception [@copelovitch2010a, @stone2011]. Empirical studies have consistently shown that geopolitical considerations play a central role in IMF lending decisions.

@@

As an international organization formally governed by a system of weighted voting and technocratic procedures, IMF is viewed as an “accountant” [@lutz2019], primarily concerned with the internal consistency of fiscal projections to guarantee timely repayment of its financial assistance. However, empirical studies suggested that countries with low international reserves, high current account deficits, and low GDP growth are more likely to receive IMF funding assistance [@barro2005; @knight1997], despite such indicators often reflecting a limited capacity to repay. This paradox leads to a widely argued statement, particularly among realists, that international institutions are “basically a reflection of the distribution of power in the world” [@mearsheimer1994]. Generally, all studies on IMF lending outcomes agree that major power impose their influence in the process. For example, studies have shown that countries strategically important to the United States are more likely to receive IMF loans, or to receive larger amounts. Specifically, this applies to countries that align with the U.S. in United Nations General Assembly and United Nations Security Council voting [@lipscy2018; @thacker1999; @dreher2009], are in geopolitical vital regions \[\], or hold significant debts to American commercial banks [@oatley2004]. This is not surprising, as the United States, as the most powerful state, has long projected its influence over international financial institutions—much like it has done with the World Bank [@wade2002; @woods2003; @clark2020]. The question this study brings about is, how this reflection of power is distributed within a given international organization, especially one that is widely perceived to be dominated by a few major powers.

The projection of great power capabilities into international organizations is sometimes uneven, because of strategic preferences, the perceived inefficiency of institutional engagement, or other possible reasons. Existing studies provide substantial evidence of U.S. influence over IMF lending decisions, they leave a crucial question open-the comparison between the U.S. and other major powers, particularly the European countries. IMF is perceived to be an institution dominated by a few major power, but the relative weight of American versus European remains ambiguous. The organization is traditionally chaired by an European national [@kahler2001],

The literature on Europe's influence in the IMF is rather fragmented and lacks systematic research. Most scholars (such as [@smaghi2004; @thacker1999 ; @miescu2014]) argue that the representatives from Europe are rather scattered, while the United States remains the dominant force. However, there are also scholars (such as [@keating2024; @lutz2014; @copelovitch2010a; @stone2004]) who argue that Europe has significant advantages in leadership, voting power, agenda setting, overseas influence, and promoting its own interests, which enables it to play an indispensable role in the issue of IMF loans.

First, European countries have structural advantages in IMF institutions. As mentioned in the introduction, the IMF's leadership has long been held by Europeans, and this gentleman's agreement between the United States and Europe has always worked in practice [@keating2024]. EU countries hold more than 30% of the voting rights in the IMF, among which Britain, France and Germany hold seats alone [@leech2005a]. Research shows that if European countries unify their seats, it is highly possible that they will have equal veto power compared to the United States [@brandner2009a].

Second, European countries dominate policy design and technical bureaucracy. [@lutz2014]'s research shows that the EU and the European Central Bank promoted the "Washington Consensus"-style austerity policy in the IMF's loans to Central and Eastern European countries (such as Latvia and Romania). This is because France and Germany have high exposure to Eastern European banking risks. Specifically, European countries have formed a consistent position within the IMF through the EURIMF (European Representatives to the IMF) and SCIMF (Sub-committee on the IMF) mechanisms, which makes it easier to put pressure on the IMF to promote policies that are beneficial to European interests (ibid.). Meanwhile, during the European debt crisis, France and Germany pushed for the participation of the IMF in rescue efforts and established a joint mechanism of the IMF, the European Commission and the European Central Bank to deal with the crisis [@veron2016]. European countries were both lenders and borrowers. However, this move has also been criticized for the IMF overly obeying the will of major European powers, and it is believed that the failure of European governments to adopt the IMF's reform suggestions is the core cause of the outbreak of the European debt crisis (ibid.).
